pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-08-04 bmc_article.tex
Updates from Overleaf

Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Additional File 1: Fig S1 for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Additional File 1: Fig S2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

We developed \textit{pipeComp} to address the need of a framework that can simultaneously evaluate the interaction of multiple tools. The work of Vieth and colleagues\cite{viethSystematic2019} already offered an important precedent in this respect, evaluating the interaction of various steps in the context of scRNAseq differential expression analysis, but did not offer a platform for doing so. Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at multiple steps. This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark. This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake} \citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Supplementary methods for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}). In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{pipeline_explanation.pdf}
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept a number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function (see Supplementary Methods for more details). \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:explanation}
\end{figure}

>>>>

Here, we develop a flexible R framework for pipeline comparison and apply it to the evaluation of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table S\ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table S\ref{tab:table1} and Additional File 1: Fig S1 for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Additional File 2). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Additional File 2 for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Additional File 1: Fig S2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

We developed \textit{pipeComp} to address the need of a framework that can simultaneously evaluate the interaction of multiple tools. The work of Vieth and colleagues\cite{viethSystematic2019} already offered an important precedent in this respect, evaluating the interaction of various steps in the context of scRNAseq differential expression analysis, but did not offer a platform for doing so. Instead, the \textit{CellBench} package was recently proposed to address a similar need \cite{suCellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at multiple steps. This avoids the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark. This, along with the set-up simplicity (demonstrated here by our separate SVA-DEA benchmark), also distinguishes \textit{pipeComp} from general workflow frameworks such as \textit{drake} \citep{landauDrakePackage2018}; indeed we estimate that the benchmarks discussed here would have taken approximately 16TB of storage had all non-redundant intermediates been saved (see the Additional File 2 for more detailed explanations of the framework and comparison to \textit{CellBench} and \textit{drake}). In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, making it easy to modify, extend with further methods, or apply to other datasets.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{pipeline_explanation.pdf}
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept a number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function (see Additional File 2 for more details). \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:explanation}
\end{figure}
